---
title: "Data preparation"
author: "Elsa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Load Data

1.1 Sample metadata
```{r}
sample_metadata <- read_tsv("data/LAHT-pilot-metadata.tsv")
```

1.2 Read Counts
```{r}
read_counts <- read_tsv("data/DMB0190_counts.tsv") %>%
    rename(genome=1)
```

### Filter read counts
```{r load_read_counts, warning=FALSE, comments="", message=FALSE, eval=FALSE}
read_counts <- read_counts %>%
  select(one_of(c("genome",sample_metadata$DNA_datasets))) %>% # individual assemblies, therefore DNA dataset = samples
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) 
```

1.3 Genome taxonomy
```{r load_genome_metadata, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_taxonomy <- read_tsv("data/genome_taxonomy.tsv") %>% 
  rename(genome = user_genome) %>% 
  separate(
    classification,
    into  = c("domain", "phylum", "class", "order", "family", "genus","species"),
    sep   = ";",
    fill  = "right",
    remove = TRUE
  ) %>%
  mutate(across(domain:species, ~ str_remove(., "^[a-z]__"))) %>% 
  select(genome,domain,phylum,class,order,family,genus,species) %>% 
  filter(genome %in% read_counts$genome)

genome_features <- read_csv("data/genomeInformation.csv") %>% 
  mutate(genome = str_remove(genome, "\\.fa$")) %>% 
  filter(genome %in% read_counts$genome)

genome_metadata <- left_join(genome_taxonomy,genome_features, by="genome")

genome_metadata <- genome_metadata %>% 
  mutate(phylum = case_when(
        phylum == "Bacillota_A" ~ "Bacillota",
        phylum == "Bacillota_C" ~ "Bacillota",
        phylum == "Bacillota_B" ~ "Bacillota",
        phylum == "Bacillota_I" ~ "Bacillota",
        TRUE ~ phylum))
```

1.4 Genome base hits (coverage)
```{r load_genome_hits, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_coverage <- read_tsv("data/DMB0190_bases.tsv") %>%
  rename(genome = 1) %>%
  filter(genome %in% read_counts$genome) %>%
  select(one_of(c("genome",sample_metadata$DNA_datasets))) 
```

1.5 Genome tree
```{r load_genome_tree, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_tree <- read_tree("data/bacteria.tree")
genome_tree$tip.label <- str_replace_all(genome_tree$tip.label,"'", "")
genome_metadata_bacteria <- genome_metadata %>% 
  filter(domain=="Bacteria")
genome_tree <- ape::keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)
```

1.6 Genome functional annotations
```{r}
genome_annotations <- read_tsv("data/gene_annotations.tsv.xz") %>%
  mutate(genome = str_extract(gene, "^[^^]+")) %>%
  mutate(ec = str_c("[EC:",ec,"]")) %>% 
  select(genome, everything())
```

1.7 Physiology data
```{r}
#resp_data <- read_tsv("data/FILENAMEHERE")
```

1.8 Immune assay data
```{r}
#pha_data <- read_tsv("data/FILENAMEHERE")
```

1.9 Behavioral data
```{r}
#activity_data <- read_tsv("data/FILENAMEHERE")
```

2. Create working objects

Transform the original data files into working objects for downstream analyses.

### Filter reads by coverage
```{r match, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# Keep only genomes and samples present in both datasets
common_genomes <- intersect(read_counts$genome, genome_coverage$genome)
common_samples <- intersect(colnames(read_counts)[-1], colnames(genome_coverage)[-1])

read_counts <- read_counts %>%
  filter(genome %in% common_genomes) %>%
  select(genome, all_of(common_samples)) %>%
  arrange(match(genome, common_genomes))

genome_coverage <- genome_coverage %>%
  filter(genome %in% common_genomes) %>%
  select(genome, all_of(common_samples)) %>%
  arrange(match(genome, common_genomes))
```

```{r filter_coverage, warning=FALSE, comments="", message=FALSE, eval=FALSE}
min_coverage=0.3

read_counts_filt <- read_counts %>%
  select(genome) %>%
  bind_cols(
    genome_coverage %>%
      select(-genome) %>% # drop genome column
      mutate(across(everything(), ~ as.numeric(. > min_coverage))) %>% # 0/1 coverage
      map2_df(.x = ., .y = read_counts %>% select(-genome), ~ .x * .y)  # multiply by read counts
  )
```

### Transform reads into genome counts
```{r calculate_genome_counts_unfiltered, warning=FALSE, comments="", message=FALSE, eval=FALSE}
readlength=150
genome_counts_filt <- read_counts_filt %>%
  slice(match(genome_metadata$genome, genome)) %>%  # reorder rows to match genome_metadata
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength)))
```

### Distill annotations into GIFTs
```{r distill_annotations, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_gifts <- distill(genome_annotations,GIFT_db,genomecol=1,annotcol=c(6,7,8,9), verbosity=TRUE)
genome_gifts <- genome_gifts[, !grepl("^S", colnames(genome_gifts))]
```

## Prepare color scheme

[AlberdiLab](www.alberdilab.dk) projects use unified color schemes developed for the [Earth Hologenome Initiative](www.earthhologenome.org), to facilitate figure interpretation.

```{r get_ehi_colors, warning=FALSE, comments="", message=FALSE, eval=FALSE}
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    mutate(phylum = str_remove(phylum, "^p__")) %>% 
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>%
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)
```

2.1 Prepare phyloseq object
```{r}
phylo_samples <- sample_metadata %>% 
  column_to_rownames("DNA_datasets") %>% 
  sample_data() #convert to phyloseq sample_data object
phylo_genome <- genome_counts_filt %>% 
  column_to_rownames("genome") %>% 
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>%
  column_to_rownames("genome") %>% 
  as.matrix() %>% 
  tax_table() #convert to phyloseq tax_table object

physeq_genome <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
physeq_genome_clr <- microbiome::transform(physeq_genome, 'clr')
```

3.0 Wrap working objects

All working objects are wrapped into a single Rdata object to facilitate downstream usage.

### Taxonomic, and other host-associated data
```{r wrap_working_objects, warning=FALSE, comments="", message=FALSE, eval=FALSE}
save(
  sample_metadata,
  genome_metadata,
  genome_metadata_bacteria,
  read_counts,
  genome_counts,
  genome_counts_filt,
  genome_tree,
  genome_gifts,
  phylum_colors,
  physeq_genome, 
  physeq_genome_clr,
  #resp_data, 
  #pha_data, 
  #activity_data,
  file = "data/data.Rdata"
)
```

### Functional data
```{r extra_objects, warning=FALSE, comments="", message=FALSE, eval=FALSE}
save(genome_gifts, 
     file = "data/genome_gifts.Rdata")
```