---
title: "Data Summary"
author: "Elsa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Load data
```{r}
load("data/data.Rdata")
```

```{r}
read_counts_row <- read_counts %>% 
  column_to_rownames(., "genome") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rename(seq=".") %>% 
  arrange(seq)
```

Number of fecal samples
```{r}
samples <- sample_metadata %>% 
  filter(sample_type == "Faecal") %>% 
  nrow() %>% 
  cat()
```

Number of MAGs
```{r}
nmags <- nrow(read_counts) %>%  # establish number of MAGs for analysis
  cat() 
```


Amount of host data (GB) - as mapped against the Calotriton arnoldi reference genome. the closest living relative to Calotriton asper with a complete reference genome
```{r}
host_GB <- sum(sample_metadata$host_data_GB)  %>% 
  cat()

metagenomic_GB <- sum(sample_metadata$metagenomic_data_GB)
```
Amount of metagenomic data (GB)
```{r}
metagenomic_GB <- sum(sample_metadata$metagenomic_data_GB) %>% 
  cat()
```
MAG Statistics

Number of new species (MAGs lacking species-level annotation)
```{r}
nonspecies <- genome_metadata %>%
    filter(species == "") %>%
    nrow()
cat(nonspecies)
```

Percentage of MAGs in catalogue lacking species-level annotation
```{r}
perct <- nonspecies*100/nmags
cat(perct)
```

Number of phyla represented in MAG catalogue
```{r}
nphyla <- genome_metadata %>%
  select(phylum) %>%
  unique() %>%
  pull() %>%
    length()
cat(nphyla)
```