---
title: "Community Composition"
author: "Elsa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Community composition

```{r load_data_community}
load("data/data.Rdata")
```

## Taxonomy overview 
change "VARIABLE" in order to compare groups of samples; i.e. experimental groups temperatures, etc... 
In this skeleton draft, the two options for VARIABLE are "A" and "B"

```{r}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(sample_metadata, by = join_by(sample == DNA_datasets)) %>%
  filter(count > 0) %>% 
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + 
    geom_bar(stat="identity", colour="white", linewidth=0.1) + 
    scale_fill_manual(values=phylum_colors) +
    facet_grid(~ water_temp, scale="free", space="free")+
    guides(fill = guide_legend(ncol = 1)) +
    theme(
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 8, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")) +
   labs(fill="Phylum",y = "Relative abundance",x="samples")
```

```{r phyla, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```
### Genus and species annotation

**Number of MAGs without species-level annotation**
```{r nonspe, comment="", message=FALSE, warning=FALSE}
genome_metadata %>%
  filter(species == "") %>%
  summarize(Mag_nospecies = n())%>%
  select(Mag_nospecies) %>% 
  pull() %>% 
  cat()
```
```{r nonspe_phylum, comment="", message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(count_total = n())
genome_metadata %>%
  filter(species == "") %>%
  group_by(phylum) %>%
  summarize(count_nospecies = n()) %>% 
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>% 
  mutate(percentage_nosp=100*count_nospecies/count_total) %>% 
  arrange(percentage_nosp) %>% 
  tt()
```

**Percentage of MAGs without species-level annotation**
```{r sp_percet, comment="", message=FALSE, warning=FALSE}
nmags <- nrow(genome_metadata)
nonspecies <- genome_metadata %>%
    filter(species == "") %>%
    nrow()
perct <- nonspecies*100/nmags
perct %>% 
  cat()
```

**Number of MAGs without genera-level annotation**
```{r nongenera, comment="", message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>%
    filter(genus == "") %>%
    nrow()
cat(nongenera)
```

**Percentage of MAGs without genera-level annotation**
```{r gen_percet, comment="", message=FALSE, warning=FALSE}
perct_g <- nongenera*100/nmags
perct_g %>% 
  cat()
```
#### Phylum Relative abundances
```{r}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome), ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == DNA_datasets)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample, phylum, water_temp) %>%
  summarise(relabun = sum(count), .groups = "drop")
```
## Alpha diversity
```{r alpha_div, comment="", message=FALSE, warning=FALSE, eval=FALSE}
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(q = 0) %>%
  as.data.frame() %>%
  dplyr::rename(richness = value)

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(q = 1) %>%
  as.data.frame() %>%
  dplyr::rename(neutral = value)

phylogenetic <- genome_counts_filt %>%
  filter(genome %in% genome_metadata_bacteria$genome) %>% 
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(q = 1, tree = genome_tree) %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = value)

alpha_div <- richness %>%
  dplyr::select(sample, richness) %>%
  full_join(neutral %>% dplyr::select(sample, neutral), by = join_by(sample == sample)) %>%
  full_join(phylogenetic %>% dplyr::select(sample, phylogenetic), by = join_by(sample == sample))
```

```{r save_alpha, eval=FALSE}
save(alpha_div, file="data/alpha_all.RData")
```

```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == DNA_datasets)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    cold_mean = mean(value[water_temp == "10"], na.rm = T),
    cold_sd = sd(value[water_temp == "10"], na.rm = T),
    hot_mean = mean(value[water_temp == "20"], na.rm = T),
    hot_sd = sd(value[water_temp == "20"], na.rm = T),
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    "Cold" = str_c(round(cold_mean, 3), "±", round(cold_sd, 3)),
    "Hot" = str_c(round(hot_mean, 3), "±", round(hot_sd, 3)),
  ) %>%
  arrange(-hot_mean) %>%
  dplyr::select(alpha, Total, "Cold", "Hot") %>%
  rename(Alpha=alpha) %>% 
  tt()
```
```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == DNA_datasets)) %>%
  mutate(water_temp = factor(water_temp, levels = c(10, 20))) %>%
  ggplot(aes(
    y = value,
    x = water_temp,
    group = water_temp,
    color = water_temp,
    fill = water_temp
  )) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(
    ~ factor(
      metric,
      levels = c("richness", "neutral", "phylogenetic"),
      labels = c("Richness", "Neutral", "Phylogenetic")
    ),
    scales = "free",
    ncol = 5
  ) +
  theme_classic()
```

```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  left_join(sample_metadata, by = join_by(sample == DNA_datasets)) %>%
  group_by(metric) %>%
  wilcox_test(value ~ water_temp) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

# Beta Div

```{r}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

 beta_q1p <- genome_counts_filt %>%
   filter(genome %in% genome_metadata_bacteria$genome) %>% 
   column_to_rownames(., "genome") %>%
   filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
   select_if(~!all(. == 0)) %>%
   hillpair(., q = 1, tree = genome_tree)

save(beta_q0n, 
     beta_q1n, 
     beta_q1p, 
     file = "data/beta.Rdata")
```

```{r}
load("data/beta.Rdata")

# set water temp as factor, not number
sample_metadata_beta <- sample_metadata %>%
  mutate(water_temp = factor(water_temp, levels = c(10, 20)))
```

```{r}
# richness 
beta_q0n_mat <- as.matrix(beta_q0n$q0S)

beta_q0n_filtered <- beta_q0n_mat[
  rownames(beta_q0n_mat) %in% sample_metadata_beta$DNA_datasets,
  colnames(beta_q0n_mat) %in% sample_metadata_beta$DNA_datasets
]

beta_q0n_dis <- as.dist(beta_q0n_filtered)

sample_metadata_q0n <- sample_metadata_beta %>%
  filter(DNA_datasets %in% labels(beta_q0n_dis)) %>%
  arrange(match(DNA_datasets, labels(beta_q0n_dis))) %>%
  column_to_rownames("DNA_datasets")

all(rownames(sample_metadata_q0n) == labels(beta_q0n_dis))

betadisper(beta_q0n_dis, sample_metadata_q0n$water_temp) %>%
  permutest(pairwise = TRUE)

adonis2(
  beta_q0n_dis ~ water_temp,
  data = sample_metadata_q0n,
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```
```{r}
# Neutral beta diversity, q = 1

beta_q1n_mat <- as.matrix(beta_q1n$q1S)

beta_q1n_filtered <- beta_q1n_mat[
  rownames(beta_q1n_mat) %in% sample_metadata_beta$DNA_datasets,
  colnames(beta_q1n_mat) %in% sample_metadata_beta$DNA_datasets
]

beta_q1n_dis <- as.dist(beta_q1n_filtered)

sample_metadata_q1n <- sample_metadata_beta %>%
  filter(DNA_datasets %in% labels(beta_q1n_dis)) %>%
  arrange(match(DNA_datasets, labels(beta_q1n_dis))) %>%
  column_to_rownames("DNA_datasets")

all(rownames(sample_metadata_q1n) == labels(beta_q1n_dis))

betadisper(beta_q1n_dis, sample_metadata_q1n$water_temp) %>%
  permutest(pairwise = TRUE)

adonis2(
  beta_q1n_dis ~ water_temp,
  data = sample_metadata_q1n,
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```

```{r}
# Phylogenetic beta diversity, q = 1

beta_q1p_mat <- as.matrix(beta_q1p$q1S)

beta_q1p_filtered <- beta_q1p_mat[
  rownames(beta_q1p_mat) %in% sample_metadata_beta$DNA_datasets,
  colnames(beta_q1p_mat) %in% sample_metadata_beta$DNA_datasets
]

beta_q1p_dis <- as.dist(beta_q1p_filtered)

sample_metadata_q1p <- sample_metadata_beta %>%
  filter(DNA_datasets %in% labels(beta_q1p_dis)) %>%
  arrange(match(DNA_datasets, labels(beta_q1p_dis))) %>%
  column_to_rownames("DNA_datasets")

all(rownames(sample_metadata_q1p) == labels(beta_q1p_dis))

betadisper(beta_q1p_dis, sample_metadata_q1p$water_temp) %>%
  permutest(pairwise = TRUE)

adonis2(
  beta_q1p_dis ~ water_temp,
  data = sample_metadata_q1p,
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```
```{r}
nmds_beta <- beta_q1n_dis %>%
  vegan::metaMDS(trymax = 500, k = 2, trace = 0) %>%
  vegan::scores() %>%
  as_tibble(rownames = "sample") %>%
  left_join(sample_metadata_beta, by = join_by(sample == DNA_datasets)) %>%
  group_by(water_temp) %>%
  mutate(
    x_cen = mean(NMDS1, na.rm = TRUE),
    y_cen = mean(NMDS2, na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r}
nmds_beta %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = water_temp)) +
  geom_point(size = 3, alpha = 0.6) +
  geom_segment(
    aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2),
    alpha = 0.4,
    show.legend = FALSE
  ) +
  scale_color_manual(
    name = "Water Temp",
    values = type_colors
  ) +
  theme_classic()
```
# Relative abundance
```{r}
# set water temp as a factor, not a number, within physeq
sample_data(physeq_genome)$water_temp <- factor(
  sample_data(physeq_genome)$water_temp,
  levels = c(10, 20)
)
```

```{r}
fit_mag_water_temp <- ancombc2(
  data = physeq_genome,
  fix_formula = "water_temp",
  group = "water_temp",
  global = FALSE,
  pairwise = FALSE,
  p_adj_method = "BH",
  lib_cut = 0,
  struc_zero = TRUE,
  alpha = 0.05
)

mag_water_temp <- fit_mag_water_temp$res

save(
  fit_mag_water_temp,
  mag_water_temp,
  file = "data/ancombc.Rdata"
)
```
```{r}
mag_enriched_20 <- mag_water_temp %>%
  filter(q_water_temp20 < 0.05) %>%
  filter(lfc_water_temp20 > 0) %>%
  arrange(desc(lfc_water_temp20))

mag_enriched_20 %>%
  nrow() %>%
  cat()

mag_enriched_20 %>%
  tt()
```
```{r}
mag_enriched_10 <- mag_water_temp %>%
  filter(q_water_temp20 < 0.05) %>%
  filter(lfc_water_temp20 < 0) %>%
  arrange(lfc_water_temp20)

mag_enriched_10 %>%
  nrow() %>%
  cat()

mag_enriched_10 %>%
  tt()
```

```{r}
## Plotting
taxonomy <- genome_metadata %>%
  distinct(genome, phylum) %>%
  rename(taxon = genome)

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate(phylum = str_replace(phylum, "[dpcofgs]__", "")) %>%
  right_join(taxonomy %>% distinct(phylum), by = "phylum") %>%
  select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80")) %>%
  distinct() %>%
  arrange(phylum)
```

```{r}
plot_data <- mag_water_temp %>%
  filter(q_water_temp20 < 0.05) %>%
  mutate(
    logFC = lfc_water_temp20,
    enriched = case_when(
      logFC > 0 ~ "20C",
      logFC < 0 ~ "10C",
      TRUE ~ NA_character_
    )
  ) %>%
  left_join(genome_metadata, by = join_by(taxon == genome))
```
```{r}
tax_color <- plot_data %>%
  distinct(phylum) %>%
  left_join(colors_alphabetic, by = "phylum") %>%
  arrange(phylum) %>%
  pull(colors, name = phylum)
```
```{r}
ggplot(plot_data, aes(x = reorder(taxon, logFC), y = logFC, fill = phylum)) +
  geom_col() +
  scale_fill_manual(values = tax_color, na.value = "grey70") +
  coord_flip() +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 6),
    axis.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14, face = "bold"),
    legend.position = "right",
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  labs(
    title = "Enriched in 10C (- LFC) vs enriched in 20C (+ LFC)",
    x = "MAGs",
    y = "Log-Fold Change",
    fill = "Phylum"
  )
```
